Relative parabolic factorizations of Chevalley groups

### 2015-05-30 dennis-vaserstein.tex
Added some introduction and stated some theorems

%Now denote by $\widetilde{\alpha}$ the maximal root of $\Phi$.
By our assumptions on $\Phi$ we have $m_s(\alpha\ssub{\mathrm{max}})=1$ hence every two roots $\alpha, \beta \in \Sigma^-_s$ 
have the same $s$-shape and
%(i.\,e. $\shape(\{s\}, \alpha = \shape(\{s\}, \beta)$).
hence by Lemma~\ref{lemma:abs} are interchanged under the action of $W(\Delta_s)$ if their length is equal. % $\alpha$ and $\beta$ if their length is equal.
%By the same argument as above we get that $W(\Delta_s)$.
Since $X_{\alpha\ssub{s}} \subseteq L_{rs}$ by the same argument as above we get $\U({\Sigma^-_s}^>, I)\subseteq L_{rs}$.
This completes the proof of the theorem for $\Phi\neq \rB_\ell$.

In the case $\Phi=\rB_\ell$ we have to show additionally that $X_{-\alpha^<_{\mathrm{max}}}(I) \leq L_{rs}$, where $\alpha^<_{\mathrm{max}}$
denotes the maximal short root of $\rB_\ell$.
By Chevalley commutator formula we have:
$$t_{-{\alpha^<_{\mathrm{max}}}}(s) = t_{-1,0}(s) = t_{1,-2}(s) \cdot {}^{t_{-2, 0}(1)}\!t_{-1,-2}(-s) \cdot t_{-1, 2}(\xi),\ s\in I.$$
On the other hand, $t_{-2,0}(1) \in \EP_s$ while we have already shown $X_{-1,-2}(I), X_{-1, 2}(I) \subseteq L_{rs}$.
This concludes the proof of the theorem.
\end{proof}

>>>>

%Now denote by $\widetilde{\alpha}$ the maximal root of $\Phi$.
By our assumptions on $\Phi$ we have $m_s(\alpha\ssub{\mathrm{max}})=1$ hence every two roots $\alpha, \beta \in \Sigma^-_s$ of the same length
have the same $s$-shape and
%(i.\,e. $\shape(\{s\}, \alpha = \shape(\{s\}, \beta)$).
hence by Lemma~\ref{lemma:abs} are interchanged under the action of $W(\Delta_s)$. % $\alpha$ and $\beta$ if their length is equal.
%By the same argument as above we get that $W(\Delta_s)$.
Since $X_{\alpha\ssub{s}} \subseteq L_{rs}$ by the same argument as above we get $\U({\Sigma^-_s}^>, I)\subseteq L_{rs}$.
This completes the proof of the theorem for $\Phi\neq \rB_\ell$.

In the case $\Phi=\rB_\ell$ we have to show additionally that $X_{-\alpha^<_{\mathrm{max}}}(I) \leq L_{rs}$, where $\alpha^<_{\mathrm{max}}$
denotes the maximal short root of $\rB_\ell$.
By Chevalley commutator formula we have:
$$t_{-{\alpha^<_{\mathrm{max}}}}(s) = t_{-1,0}(s) = t_{1,-2}(s) \cdot t_{-1,-2}(-s)^{t_{-2, 0}(-1)} \cdot t_{-1, 2}(\xi),\ s\in I.$$
On the other hand, $t_{-2,0}(1) \in \EP_s$ while we have already shown $X_{-1,-2}(I), X_{-1, 2}(I) \subseteq L_{rs}$.
This concludes the proof of the theorem.
\end{proof}

### 2015-05-30 paper.tex
Added some introduction and stated some theorems

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
%TODO: Deprecate Um -- Use Ums/Umd whenever possible
\DeclareMathOperator{\Um}{Um}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

\section{Introduction}\label{sec:intro}

>>>>

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
%TODO: Deprecate Um -- Use Ums/Umd whenever possible
\DeclareMathOperator{\Um}{Um}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

\section{Introduction}\label{sec:intro}
Let $\Phi$ be a reduced irreducible root system.

\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that
\begin{itemize} \item either $\sr(I) = 1$;
 \item or $R = p^{-1}\cdot \mathbb{Z}$ for some prime $p$. \end{itemize}
In the latter case assume additionally that Generalized Riemann Hypothesis holds.
Then for any $\Phi$ the length of the congruence subgroup $\G(\Phi, R, I)$ does not exceed $C_1|\Phi| + C_2$. \end{thm}
%TODO: Write explicit estimate for C_1 and C_2.
The theorem above generalizes one of the results of \cite{VavSmSuUnitrEng} where such bounds were obtained in the absolute case (i.e. when $I=R$).

The next our main result concerns subsystem factorizations of Chevalley groups formulated in terms of submaximal subgroups of type $\rA_\ell$.
The case when $R=I$ is a field this question was studied by N.~Nikolov in \cite{NikProdDecomp}. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $n$ over a finite field.
Then $G$ equals the product of at most $200$ conjugates of $\SL_n$. \end{thm*}

It is also classically known that over arbitrary field $F$ the group $\SL(n+1, F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, from the classical Dennis---Vaserstein theorem (see~\cite{Va69} and \cite[Lemma~2.1]{ST76})
it follows that $\SL_{n+1}(R, I)$ is a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leq n-1$.
Using another Dennis---Vaserstein type factorization we obtain the following result.
\begin{thm} Assume that $\sr(I) = 2$ then the group $\Spin_{2\ell}(R, I)=\G_{sc}(\rD_\ell, R, I)$ is a product of at most $C$ copies of subgroups of type $\rA_{\ell-1}$. \end{thm}

Another applications of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{NikProdDecomp}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies similar factorization with $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies similar factorization for the elementary subgroup $\E(\Phi, R)$ with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.

\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that $\sr(R)\leq n$, where
\begin{itemize}
 \item $n=1$ and $\Phi$ is either classical or of type $\rE_\ell$;
 \item $n=2, 3$ and $\Phi$ is simply laced.
\end{itemize}
Then $\G(\Phi, R, I)$ is a product of at most $q_{n}(\Phi)$ copies of $\SL_{n+1}(R, I)$, where each $q_{n}(\Phi)$, $n=1,2,3$ is a quadratic function of $\rk(\Phi)$ given in Table~XXX.
\end{thm}
